raptr: Representative and Adequate Prioritisation Toolkit in R

### 2016-11-18 code/rmarkdown/article.Rmd
added in Richs, Hughs, and Jonathan's comments

---
title: "raptr: Representative and Adequate Prioritisation Toolkit in R"
author: |
  | Jeffrey O. Hanson$^1$, Jonathan R. Rhodes$^2$, Hugh P. Possingham$^1$, Richard A. Fuller$^1$
  | $^1$School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  | $^2$School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
  | Correspondance should be addressed to jeffrey.hanson@uqconnect.edu.au
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: ["conservation", "biodiversity", "MILP", "reserve-selection", "optimisation", "protected area"]
abstract: "

1. An underlying aim in conservation planning is to maximise the long-term persistence of biodiversity. To fulfill this aim, the ecological and evolutionary processes that sustain biodiversity must be preserved. One way to conserve such processes at the feature level (eg. species, ecosystem) is to preserve a representative sample of the physical attributes that underpin them across the feature's geographic distribution. For example, preserving the adaptive processes currently acting on a species might be crucial for its long-term persistence in a world where environmental change is accelerating. By conserving individuals with the ability to persist in a range of climatic conditions--physical attributes associated with adaptation--protected areas can foster evolutionary processes. Yet current approaches overwhelmingly focus on securing a target proportion of each features' geographic range; little attention is directed towards targeting a representative sample of them.

2. To address this issue, we developed the \\texttt{raptr: Representative and Adequate Prioritisation Toolkit in R} to guide reserve selection using explicit targets for representing biodiversity processes as well as more typical area protection targets. Users set \"amount targets\"--similar to conventional reserve selection methods--to secure a target proportion of the feature's geographic distribution. Additionally, users set \"space targets\" to ensure that solutions also secure a representative sample of an evolutionary or ecologically relevant attribute (space) across a feature's geographic distribution (eg. an attribute expressing variation in genetic characteristics between individuals, or variation in climatic conditions between areas). We demonstrate the functionality of this package using simulations and two case studies. We generate solutions that secure a proportion of the species' distributions using amount targets--representing conventional reserve selection--and compare them with solutions generated using both amount and space targets.

3. We show that markedly different solutions emerge when explicitly considering within feature representation in reserve selection. Our simulations suggest that including representativeness in conservation planning is important where biodiversity features have multimodal distributions in an attribute space. The case studies show that explicitly setting space-based targets can result solutions that preserve a much more representative sample of features. Additionally, the case studies suggest that securing a representative sample of features may not require a much  greater area of reservation than traditional approaches.

4. The \\texttt{raptr R} package provides a unified framework for achieving spatial conservation prioritisations that secure an adequate and representative sample of features. Prioritisations that fulfill that , resulting in a greater chance of achieving long-term biodiversity persistence than area-based planning alone.

@@

Perhaps the most fundamental aim of conservation is to maximise the long-term persistence of biodiversity [@r423; @r255]. To achieve this, conservation actions must preserve biodiversity patterns (eg. populations, species, ecosystems), but also crucially the processes that sustain them. One of the major tangible achievements of modern conservation has been the act of setting aside areas for preservation [@r440]. Reserve networks buffer species from gross threatening processes and set the stage for enhanced management interventions [@r465]. Since the resources available for conservation action are limited, protected area networks must be sited in places that satisfy conservation objectives for minimal cost [@r255].

Today, the most widely used conservation planning tools focus on biodiversity patterns [\texttt{Zonation} and \texttt{Marxan}; @r429; @r22]. These tools allow decision makers to obtain solutions that secure a proportion of the geographic range of biological features (populations, species, or ecosystems of conservation interest). One method to incorporate data on the ecological and environmental processes that sustain features is to conserve a representative sample of the physical attributes that underpin these processes. To achieve this, current methods typically involve partitioning each feature into sub-groups based on an attribute variable that relates to a biodiversity process of interest [@r9; @r247; @r448]. For instance, by dividing species distributions into sub-groups according to habitat discontinuities, and ensuring that each sub-group is represented in the solution, conservation planners can obtain prioritisations that promote adaptive processes [@r2]. However, using this method is challenging because biodiversity data are often continuous and multi-dimensional.

Simply partitioning features based on a continuous variables and setting targets for each sub-group may not necessarily result in a solution that secures a representative sample of the continuous variable. For instance, if a feature was split into four sub-features based on a single continuous variable, and a solution preserve 10 % of the geographic range of each sub-feature, how much of the original continuous variation does the solution represent? The answer is not immediately clear. If the solution sites reserves evenly along the original variable, then the solution may sample far more than 10 % of the original variation. Whereas, if the solution sites reserves near the values used to divide the continuous values [as observed when partitioning continuous geographic variation based on political boundaries; @r483], then it may sample much less of the original variation. Conventional reserve selection problems cannot differentiate between either of these solutions--they are both optimal--though the poorer solution may be favored when using additional constraints to reduce fragmentation. This limitation has not arisen recently. In fact, it dates back to some of the earliest reserve selection methods [eg. @r295].

During the early 1990s, Faith and Walker [-@r480] developed a decision support tool (\texttt{DIVERSITY}) to identify prioritisations that secure a representative sample of continuous environmental variation across a study area. Although this method was originally pitched as an alternative to species-based conservation planning [@r481; @r218; @r203], it can also identify solutions that conserve a representative sample of variation across a single species' range. Recent work has built on this reserve selection method by solving problems with more advanced optimisation routines [@r482]. However, \texttt{DIVERSITY}-based reserve selection methods are not widely used today, in part, because they do not accommodate explicit amount-based targets for multiple features (_cf_. \texttt{Marxan} and \texttt{Zonation}).

One of the key issues in reserve selection is the lack of a unified decision support tool that uses explicit and separate targets to obtain solutions that secure an adequate amount of habitat for each feature and a representative sample of each feature. To begin to fill this gap, here we unite the ideas underpinning \texttt{DIVERSITY} and \texttt{Marxan} into new formulations of the reserve selection problem, and implement them in the \texttt{raptr R} package. This \texttt{R} package provides decision makers with the tools to generate prioritisations based on data that relate to biodiversity patterns and processes. Here, we aim to the provide an in-depth understanding of this \texttt{R} package and explore its functionality.

# Methods
## OPTIMISATION PROBLEM FORMULATION
Biodiversity features are defined as the entities that the prioritisation is required to preserve (eg. species, ecosystems). Spatial attributes are defined as the variation across the feature' range that the prioritisation is required to sample. They can be intrinsic variation (eg. genetic or trait) variation that is extrinsic to the feature (eg. environmental conditions). There should, however, be a reasonable underlying hypothesis that relates this variation to the biodiversity processes that the decision maker aims to preserve.

A set of attributes is conceptualised as a space (termed an "attribute space"). Each planning unit may be thought to exist as a single point inside a given space (). For example, a decision maker may require a prioritisation that represents populations along climatic gradients. To achieve this, the decision maker might use an "climatic" attribute space with dimensions relating to mean annual temperature ($^{\circ}$C) and precipitation (mm). Any given combination of temperature and precipitation may be conceived as a point in this environmental space. By associating planning units with climatic data, they can be mapped from geographic space to this environmental attribute space. In addition to planning units, attribute spaces also contain demand points.

@@

The formulations used to express the reserve selection problem in the \texttt{raptr R} package are based on elements of the \texttt{Marxan}, \texttt{DIVERSITY}, and the uncapacitated facility location problems [@r451]. Although the \texttt{raptr R} package utlises two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in Appendix S1. These formulations are based on unreliable and reliable facility location problems and have been named accordingly [@r16]. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied by features when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the set.

@@

## OPTIMISATION
The unreliable formulation is non-linear. However, the non-linear components can be linearised using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearised using methods described by Beyer et al. [-@r426]. Linearised versions of the problems can be solved using commercial exact algorithm solvers. The \texttt{raptr} R package provides functions to express conservation planning data as optimisation problems using linearised versions of the unreliable and reliable formulations (see Appendix S1 for formulation). These optimisation problems can then be solved to generate prioritisations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website.

@@

# Implications and future directions
The \texttt{raptr} R package provides a unified approach to reserve selection. One of the key advantages of this package is that it is general enough to incorporate any spatial variable as an attribute space. This package can accommodate intrinsic or extrinsic variation to the feature(s). For example, adaptation processes could be secured using environmental variation or genetic variation. Ecological processes, such as predator-prey interactions, could be secured by capturing intra-specific trophic variation. Additionally, different attribute spaces with different targets can be specified for different species. As long as the variation can be described using Euclidean distances, the \texttt{R} package can be used to obtain a representative sample (note that data may require transformation or dimension reduction to meet this assumption). This \texttt{R} package provides decision makers with the tools to generate prioritisations that secure both biodiversity patterns and the processes that maintain biodiversity. Additionally, the package contains functionality to accommodate uncertainty in the distribution of features, and also identify suitably connected reserves. Both the simulated and case study species suggest that explicitly considering proxies for biodiversity processes can significantly alter prioritisations.

The degree to which a prioritisation truly secures a representative sample of a feature depends on the attribute spaces and distribution of demand points chosen by the decision maker. Ultimately, the space-based targets are set as a proportion based on the distribution of the demand points. As a consequence, if the decision maker uses an inappropriate set of spatial variables to construct an attribute space, or an inappropriate set of demand points , then the optimal solution will not be effective. We therefore stress that decision makers must carefully consider which biodiversity processes need to be reflected in the prioritisation, and which spatial data can be used to represent these processes. Note that maximising one process can be detrimental to another. For instance, to maximise the geographic spread of a prioritisation, reserves need to be further away from each other, yet to maximise connectivity, reserves may need to be closer together. To assist in the selection of appropriate demand points, the \texttt{R} package provides several routines for generating demand points (see the `make.DemandPoints` function). These routines use kernels to automatically generate demand points based on the distribution of features in spatial data sets [using @r231; @r450].

The formulation requires spatially comprehensive data to map attribute spaces to planning units. For instance, in both case-studies, all planning units occupied by the species were associated with coordinates in the attribute spaces. However, most real-world data sets are patchy--some planning units will be occupied by species for which attribute space data is not available. To use such patchy data with the \texttt{raptr R} package, the gaps in the data must first be filled in. Spatially explicit models [eg. krigging or generalised dissimilarity models; @r467; @r466] could be used to estimate data for locations that are missing from the data set as a pre-processing step. This approach has been successfully applied to a range of biological data sets [eg. @r468].

To maximise the long-term persistence of biodiversity, decision makers need to identify prioritisations that preserve existing patterns of biodiversity and the processes that support them. To achieve this, conservation planners need a decision support tool that can explicitly accommodate data on biodiversity patterns and processes. Here, we developed the \texttt{raptr R} package to fill this void. By exploring the functionality of this package using several simulated and case study species, we found that including space-based targets can result in substantially different solutions.

# Acknowledgements
JOH is funded by an Australian Postgraduate Award (APA) scholarship. RAF has an Australian Research Council Future Fellowship. This work was supported by the Centre for Biodiversity and Conservation Science (CBCS).

>>>>

---
title: "raptr: Representative and Adequate Prioritisation Toolkit in R"
author: |
  | Jeffrey O. Hanson$^1$, Jonathan R. Rhodes$^2$, Hugh P. Possingham$^{1,3}$, Richard A. Fuller$^1$
  | $^1$School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  | $^2$School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
  | $^3$The Nature Conservancy, South Brisbane, QLD, Australia
  | Correspondance should be addressed to jeffrey.hanson@uqconnect.edu.au
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: ["conservation", "biodiversity", "MILP", "reserve-selection", "optimisation", "protected area"]
abstract: "

1. An underlying aim in conservation planning is to maximise the long-term persistence of biodiversity. To fulfill this aim, the ecological and evolutionary processes that sustain biodiversity must be preserved. One way to conserve such processes at the feature level (eg. species, ecosystem) is to preserve a representative sample of the physical attributes that underpin them across the feature's geographic distribution. For example, preserving the adaptive processes currently acting on a species might be crucial for its long-term persistence in a world where environmental change is accelerating. By conserving individuals with the ability to persist in a range of climatic conditions--physical attributes associated with adaptation--protected areas can foster evolutionary processes. Despite this, current approaches overwhelmingly focus on securing a target proportion of each features' geographic range and little attention has been directed towards targeting a representative sample of them.

2. To address this issue, we developed the \\texttt{raptr: Representative and Adequate Prioritisation Toolkit in R} to guide reserve selection using explicit targets for representing spatially explicit variables that underpin biodiversity processes as well as more typical area protection targets. Users set \"amount targets\"--similar to conventional reserve selection methods--to secure a target proportion of the feature's geographic distribution. Additionally, users set attribute \"space targets\" to ensure that reserve systems also secure a representative sample of evolutionary or ecologically relevant attributes (space) across a feature's geographic distribution (eg. an attribute space expressing variation in genetic characteristics between individuals, or variation in climatic conditions between areas). We demonstrate the functionality of this package using simulations and two case studies. We generate solutions that secure a proportion of the species' distributions using amount targets--representing conventional reserve selection--and compare them with solutions generated using both amount and attribute space targets.

3. We show that markedly different solutions emerge when explicitly considering within feature representation in reserve selection. Our simulations suggest that including representativeness in conservation planning is important where biodiversity features have multimodal distributions in an attribute space. The case studies show that explicitly setting space-based targets can result in solutions that preserve a much more representative sample of features. Additionally, the case studies suggest that securing a representative sample of features may not require a much  greater area of reservation than traditional approaches.

4. The \\texttt{raptr R} package provides a unified framework for achieving spatial conservation prioritisations that secure an adequate and representative sample of features. Prioritisations that achieve this are likely to result in a greater chance of achieving long-term biodiversity persistence than area-based planning alone.

@@

Perhaps the most fundamental aim of conservation is to maximise the long-term persistence of biodiversity [@r423; @r255]. To achieve this, conservation actions must preserve biodiversity patterns (eg. populations, species, ecosystems), but also crucially the processes that sustain them. One of the major tangible achievements of modern conservation has been the act of setting aside areas for preservation [@r484; @r440]. Reserve networks buffer species from gross threatening processes and set the stage for enhanced management interventions [@r465]. Since the resources available for conservation action are limited, protected area networks must be sited in places that satisfy conservation objectives for minimal cost [@r255].

Today, the most widely used conservation planning tools focus on biodiversity patterns [\texttt{Zonation} and \texttt{Marxan}; @r429; @r22]. Decision makers can use these tools to obtain solutions that secure a proportion of the geographic range of each biological feature (populations, species, or ecosystems) of conservation by setting targets. One method to incorporate data on the ecological and environmental processes that sustain biological features is to conserve a representative sample of the physical attributes that underpin these processes. To achieve this, current methods typically involve partitioning each feature into sub-groups based on an attribute variable that relates to a biodiversity process of interest [@r9; @r247; @r448]. For instance, by dividing species distributions into sub-groups according to habitat discontinuities, and ensuring that each sub-group is represented in the solution, conservation planners can obtain prioritisations that promote adaptive processes [@r2]. However, using this method is challenging because biodiversity data are often continuous and multi-dimensional.

Simply partitioning features based on continuous variables and setting targets for each sub-group may not necessarily result in a solution that secures a representative sample of the continuous variable. For instance, if a feature was split into four sub-features based on a single continuous variable, and a solution preserve 10 % of the geographic range of each sub-feature, how much of the original continuous variation does the solution represent? The answer is not immediately clear. If the solution sites reserves evenly along the original variable, then the solution may sample far more than 10 % of the original variation. Whereas, if the solution sites reserves near the values used to divide the continuous values [as observed when partitioning continuous geographic variation based on political boundaries; @r483], then it might sample much less of the original variation. Conventional reserve selection problems cannot differentiate between these solutions--they are both optimal--though the poorer solution may be favored when using additional constraints to reduce cost or fragmentation. This limitation has been known of quite some time. In fact, it dates back to some of the earliest reserve selection methods [eg. @r295].

During the early 1990s, Faith and Walker [-@r480] developed a decision support tool (\texttt{DIVERSITY}) to identify prioritisations that secure a representative sample of continuous environmental variation across a study area. Although this method was originally pitched as an alternative to species-based conservation planning [@r481; @r218; @r203], it can also identify solutions that conserve a representative sample of variation across a single species' range. Recent work has built on this reserve selection method by solving problems with more advanced optimisation algorithms [@r482]. However, \texttt{DIVERSITY} has limited utility for species-based conservation planning, in part, because it cannot accommodate multiple biological features. Additionally, unlike \texttt{Marxan} and \texttt{Zonation}, users cannot set intuitive targets to obtain cost effective solutions.

Conservation planners lack a decision support tool that lets them set explicit targets to obtain solutions that (1) secure an adequate amount of habitat for each feature and (2) a representative sample of each feature. To begin to fill this gap, here we unite the ideas underpinning \texttt{DIVERSITY} and \texttt{Marxan} into new formulations of the reserve selection problem, and implement them in the \texttt{raptr R} package. This \texttt{R} package provides decision makers with the tools to generate prioritisations based on data that relate to biodiversity patterns and processes. Here, we aim to the provide an in-depth understanding of this \texttt{R} package and explore its functionality.

# Methods
## PROBLEM FORMULATION
Biodiversity features are defined as the entities that the prioritisation is required to preserve (eg. species, ecosystems). Spatial attributes are defined as the variation across the feature' range that the prioritisation is required to sample. They can be intrinsic variation (eg. genetic or trait) variation that is extrinsic to the feature (eg. environmental conditions). There should, however, be a reasonable underlying hypothesis that relates this variation to the biodiversity processes that the decision maker aims to preserve.

A set of attributes is conceptualised as a space (termed an "attribute space"). For example, a decision maker may require a prioritisation that represents populations along climatic gradients. To achieve this, the decision maker might use a "climatic" attribute space with dimensions relating to mean annual temperature ($^{\circ}$C) and precipitation (mm). Any given combination of temperature and precipitation may be conceived as a point in this environmental space. Although they exist as polygons, for simplicity, each planning unit may be thought to exist as a single point inside a given space. By associating the planning units with climatic data, and calculating a representative statistic for each planning unit (eg. mean), they can be mapped from geographic space to this environmental attribute space. In addition to planning units, attribute spaces also contain demand points.

@@

The formulations used to express the reserve selection problem in the \texttt{raptr R} package are based on elements of the \texttt{Marxan}, \texttt{DIVERSITY}, and uncapacitated facility location problems [@r451]. Although the \texttt{raptr R} package utlises two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in Appendix S1. These formulations are based on unreliable and reliable facility location problems and have been named accordingly [@r16]. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied by features when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the set.

@@

## OPTIMISATION
The unreliable formulation is non-linear. However, the non-linear components can be linearised using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearised using methods described by Beyer et al. [-@r426]. Linearised versions of the problems can be solved using commercial exact algorithms. The \texttt{raptr} R package provides functions to express conservation planning data as optimisation problems using linearised versions of the unreliable and reliable formulations (see Appendix S1 for formulation). These optimisation problems can then be solved to generate prioritisations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website.

@@

# Implications and future directions
The \texttt{raptr} R package provides a unified approach to reserve selection. This \texttt{R} package provides decision makers with the tools to generate prioritisations that secure intra-specific as well as as inter-specific biodiversity patterns. Additionally, the package contains functionality to accommodate uncertainty in the distribution of features, and also identify suitably connected reserves. Both the simulated and case study species suggest that explicitly considering representation as well as amount targets can significantly alter prioritisations.

One of the key advantages of this package is that it is general enough to incorporate any spatially explicit variable as an attribute space. This package can accommodate intrinsic or extrinsic variation in the feature(s). For example, adaptation processes could be secured using environmental variation [eg. @r247]. Predator-prey processes could be conserved by capturing the overlapping distributions of predators and each of their prey species [eg. @r485]. Additionally, space targets can be used to achieve representation for reasons that are not related to biodiversity processes, such as obtaining a geographically representative set of reserve so there is equitable access to parks by people [eg. @r483]. As long as the variation can be described using Euclidean distances--which could be achieved through transformation or dimension reduction techniques--the \texttt{R} package can be used to obtain a representative sample.

The degree to which a prioritisation truly secures a representative sample of a feature depends on the attribute spaces and the distribution of demand points chosen by the decision maker. Ultimately, the space-based targets are set as a proportion based on the distribution of the demand points. As a consequence, if the planner uses an inappropriate set of spatial variables to construct an attribute space, or an inappropriate set of demand points, then the optimal solution will not effectively represent the decision maker's objectives. We therefore stress that planners must carefully consider which biodiversity processes need to be reflected in the prioritisation, and which spatial data can be used to represent these processes. Note that maximising for one attribute space can result in trading off against another. For instance, to maximise the geographic spread of a prioritisation, reserves need to be further away from each other, yet to maximise connectivity, reserves may need to be closer together. Demand points should be distributed across the full range of variation in an attribute space to obtain solutions that secure a representative sample of the attribute space. To assist the generation of appropriate demand points, the \texttt{R} package provides several routines for generating demand points (see the `make.DemandPoints` function). These routines use kernels to automatically generate demand points based on the distribution of features in spatial data sets [using @r231; @r450].

The formulation requires spatially comprehensive data to map attribute spaces to planning units. However, many real-world data sets are patchy. For example, due to constrained resources, genetic data may have only been collected in some planning units. To use such patchy data with the \texttt{raptr R} package, the planning units that are missing data must either be omitted (as was done in the second case-study), or their missing values must be estimated. Missing data could be estimated using spatially explicit models [eg. krigging or generalised dissimilarity models; @r467; @r466]. This approach has been successfully applied to a range of biological data sets [eg. @r468].

To maximise the long-term persistence of biodiversity, decision makers need to identify prioritisations that preserve existing patterns of biodiversity and the processes that support them. To achieve this, conservation planners need a decision support tool that can deliver cost-effective prioritisations that secure an adequate amount of a representative sample of biodiversity features. Here, we developed the \texttt{raptr R} package to fill this void. By exploring the functionality of this package using several simulated and case study species, we found that including space-based targets can result in substantially different solutions.

# Acknowledgements
JOH is funded by an Australian Postgraduate Award (APA) scholarship. RAF and HPP have Australian Research Council Fellowships. This work was supported by the Centre for Biodiversity and Conservation Science (CBCS).
